Towards Proof-Reconstruction of Problems in Classical Propositional Logic

### 2017-09-28 paper/sections/metis-language-and-proofs.tex
[ paper ] working on normalization subsubsection.

For instance, let us express the problem
$p\, \vdash \neg (p \wedge \neg p) \vee (q \wedge \neg q)$
in \TPTP syntax as follows, declaring an \tptpaxiom $p$, and the
expected conclusion, the  \tptpconjecture
\verb!~ ((p & ~ p) | (q & ~ q))!.

@@

\begin{figure}
\[
% \scalebox{0.9}{
\begin{bprooftree}
  \AxiomC{}
  \RightLabel{axiom}
  \UnaryInfC{$C$}
\end{bprooftree}
\qquad
\begin{bprooftree}
  \AxiomC{}
  \RightLabel{assume $L$}
  \UnaryInfC{$L \vee \neg L$}
\end{bprooftree}
\qquad
\begin{bprooftree}
  \AxiomC{}
  \RightLabel{refl $t$}
  \UnaryInfC{$t = t$}
\end{bprooftree}
\qquad
\begin{bprooftree}
  \AxiomC{$C$}
  \RightLabel{subst $\sigma$}
  \UnaryInfC{$\sigma\,C$}
\end{bprooftree}
\qquad
% }
\]
\[
% \scalebox{0.9}{
\begin{bprooftree}
  \AxiomC{}
  \RightLabel{equality $L$ $p$ $t$}
  \UnaryInfC{$\neg (L[p] = t) \vee \neg L \vee L[ p \mapsto t]$}
\end{bprooftree}
\qquad
\begin{bprooftree}
  \AxiomC{$L \vee C$}
  \AxiomC{$\neg L \vee D$}
  \RightLabel{resolve $L$}
  \BinaryInfC{$C \vee D$}
\end{bprooftree}
% }
\]
\caption{Inference rules of the \texttt{Metis} prover.}
\label{fig:metis-inferences}
\end{figure}

These proofs are directed acyclic graphs, trees of refutations. Each
node stands for an application of an inference rule and the leaves
in the tree represent formulas in the given problem. Each node is
labeled with a name of the inference rule (e.g., \canonicalize in
Fig.~\ref{fig:metis-example}). Each edge links a premise with one
conclusion. The proof graphs have at their root the conclusion
$\bot$, since \Metis delivers proofs by refutation.

@@

\begin{figure}
\centering
\begin{bprooftree}\tt
  \AxiomC{}
  \RightLabel{\footnotesize negate}
  \UnaryInfC{$\neg p$}
  \RightLabel{\footnotesize strip}
  \UnaryInfC{$\neg p$}
  \AxiomC{}
  \RightLabel{\footnotesize axiom}
  \UnaryInfC{$p$}
  \RightLabel{\footnotesize canonicalize}
  \UnaryInfC{$p$}
  \RightLabel{\footnotesize simplify}
  \BinaryInfC{$\bot$}
  \RightLabel{\footnotesize canonicalize}
  \UnaryInfC{$\bot$}
\end{bprooftree}
\caption{The \Metis' proof tree for $p \vdash p$ from the
derivation in Fig.~\ref{fig:metis-proof-tstp}}
\label{fig:metis-example}
\end{figure}

@@

\begin{figure}
\[%\scalebox{0.9}{
  \begin{bprooftree}
    \AxiomC{$\varphi \wedge \bot$}
    \UnaryInfC{$\bot$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$\varphi \wedge \top$}
    \UnaryInfC{$\varphi$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$\varphi \wedge \neg \varphi$}
    \UnaryInfC{$\bot$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$\varphi \wedge \varphi$}
    \UnaryInfC{$\varphi$}
  \end{bprooftree}
%}
\]

\[%\scalebox{0.9}{
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \bot$}
    \UnaryInfC{$\varphi$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \top$}
    \UnaryInfC{$\top$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \neg \varphi$}
    \UnaryInfC{$\top$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \varphi$}
    \UnaryInfC{$\varphi$}
  \end{bprooftree}
%}
\]
\caption{Definitions for conjunction and disjunction simplification
in the \canonicalize inference.}
\label{fig:conjunctive-disjunctive-simpl}
\end{figure}

@@

Moreover, we are not completely covered every detail in the behavior
for all \Metis' inference rules. It is somewhat complicated to grasp
completely the inference rules since the source code is around the
10.000 lines of code.

>>>>

For instance, let us express the problem
$p\, \vdash \neg (p \wedge \neg p) ∨ (q \wedge \neg q)$
in \TPTP syntax as follows, declaring an \tptpaxiom $p$, and the
expected conclusion, the  \tptpconjecture
\verb!~ ((p & ~ p) | (q & ~ q))!.

@@

\begin{figure}
\[
% \scalebox{0.9}{
\begin{bprooftree}
  \AxiomC{}
  \RightLabel{axiom}
  \UnaryInfC{$C$}
\end{bprooftree}
\qquad
\begin{bprooftree}
  \AxiomC{}
  \RightLabel{assume $L$}
  \UnaryInfC{$L ∨ \neg L$}
\end{bprooftree}
\qquad
\begin{bprooftree}
  \AxiomC{}
  \RightLabel{refl $t$}
  \UnaryInfC{$t = t$}
\end{bprooftree}
\qquad
\begin{bprooftree}
  \AxiomC{$C$}
  \RightLabel{subst $\sigma$}
  \UnaryInfC{$\sigma\,C$}
\end{bprooftree}
\qquad
% }
\]
\[
% \scalebox{0.9}{
\begin{bprooftree}
  \AxiomC{}
  \RightLabel{equality $L$ $p$ $t$}
  \UnaryInfC{$\neg (L[p] = t) ∨ \neg L ∨ L[ p \mapsto t]$}
\end{bprooftree}
\qquad
\begin{bprooftree}
  \AxiomC{$L ∨ C$}
  \AxiomC{$\neg L ∨ D$}
  \RightLabel{resolve $L$}
  \BinaryInfC{$C ∨ D$}
\end{bprooftree}
% }
\]
\caption{Inference rules of the \texttt{Metis} prover.}
\label{fig:metis-inferences}
\end{figure}

These proofs are directed acyclic graphs, trees of refutations. Each
node stands for an application of an inference rule and the leaves
in the tree represent formulas in the given problem. Each node is
labeled with a name of the inference rule (e.g., \canonicalize in
Fig.~\ref{fig:metis-example}). Each edge links a premise with one
conclusion. The proof graphs have at their root the conclusion
$⊥$, since \Metis delivers proofs by refutation.

@@

\begin{figure}
\centering
\begin{bprooftree}\tt
  \AxiomC{}
  \RightLabel{\footnotesize negate}
  \UnaryInfC{$\neg p$}
  \RightLabel{\footnotesize strip}
  \UnaryInfC{$\neg p$}
  \AxiomC{}
  \RightLabel{\footnotesize axiom}
  \UnaryInfC{$p$}
  \RightLabel{\footnotesize canonicalize}
  \UnaryInfC{$p$}
  \RightLabel{\footnotesize simplify}
  \BinaryInfC{$⊥$}
  \RightLabel{\footnotesize canonicalize}
  \UnaryInfC{$⊥$}
\end{bprooftree}
\caption{The \Metis' proof tree for $p \vdash p$ from the
derivation in Fig.~\ref{fig:metis-proof-tstp}}
\label{fig:metis-example}
\end{figure}

@@

\begin{figure}
\[%\scalebox{0.9}{
  \begin{bprooftree}
    \AxiomC{$φ ∧ ⊥$}
    \UnaryInfC{$⊥$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$φ ∧ ⊤$}
    \UnaryInfC{$φ$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$φ ∧ ¬ φ$}
    \UnaryInfC{$⊥$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$φ ∧ φ$}
    \UnaryInfC{$φ$}
  \end{bprooftree}
%}
\]

\[%\scalebox{0.9}{
  \begin{bprooftree}
    \AxiomC{$φ ∨ ⊥$}
    \UnaryInfC{$φ$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$φ ∨ ⊤$}
    \UnaryInfC{$⊤$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$φ ∨ ¬ φ$}
    \UnaryInfC{$⊤$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$φ ∨ φ$}
    \UnaryInfC{$φ$}
  \end{bprooftree}
%}
\]
\caption{Definitions for conjunction and disjunction simplification
in the \canonicalize inference.}
\label{fig:conjunctive-disjunctive-simpl}
\end{figure}

@@

### 2017-09-28 paper/sections/proof-reconstruction.tex
[ paper ] working on normalization subsubsection.

\begin{lemma}[lem-conjunct$_{∨}$]
  \label{lem:lem-conjunct-or}
  $Γ ⊢ φ$, and $ψ \in\Prop$ then $Γ ⊢ conjunct_{∨}(φ, ψ)$, where,
  \begin{align*}
  \begin{split}
    \label{eq:conjunct-or-definition}
    conjunct_{∨}(φ, ψ) =
    \begin{cases}
      ψ, &\text{if }reorder_{∨}(φ, ψ) ≡ ψ\\
      ψ, &\text{if }ψ ≡ ψ₁ ∧ ψ₂, conjunct_{∨}(φ, ψ₁) = ψ₁\\
        &\text{and } reorder_{∨}(φ, ψ₂) ≡ ψ₂\\
      ψ, &\text{if }φ ≡ φ₁ ∧ φ₂, conjunct_{∨}(φ₁, ψ) ≡ ψ\\
      ψ, &\text{if }φ ≡ φ₁ ∧ φ₂, conjunct_{∨}(φ₂, ψ) ≡ ψ\\
      φ, &\text{otherwise.}
    \end{cases}
  \end{split}
  \end{align*}
\end{lemma}

@@

In a disjunction, $φ = φ₁ ∨ φ₂ ∨ \cdots ∨ φₙ$, we say $ψ ∈_{∨} φ$,
if there is some $i = 1, \cdots, n$ such that $ψ = φᵢ$.
Then, we define the $rm_{∨}$ function to remove in a disjunction φ, those propositions ψ repeated in the formula.

@@

\begin{equation*}
\label{eq:rm-and-or}
  rm_{∧∨}(φ) =
  \begin{cases}
    rm_{∧∨}(φ₁) ∧ rm_{∧∨}(φ₂), &\text{if }φ ≡ φ₁ ∧ φ₂\\
    rm_{∨}(rconj(φ)),  &\text{otherwise.}
  \end{cases}
\end{equation*}

@@

Finally, we remove redundancies in a proposition in its conjunctive normal form, by composing the functions described above in the Eq.~\ref{eq:rm-cnf} for the function $cleanCNF$.

\begin{equation*}
  \label{eq:rm-cnf}
  cleanCNF = rm_{∧} ∘ rm_{∧∨} ∘ rconj ∘ cnf.
\end{equation*}

>>>>

\begin{lemma}[lem-conjunct$_{∨}$]
  \label{lem:lem-conjunct-or}
  $Γ ⊢ φ$, and $ψ \in\Prop$ then $Γ ⊢ conjunct_{∨}(φ, ψ)$, where,
  \begin{align*}
  \begin{split}
    \label{eq:conjunct-or-definition}
    conjunct_{∨}(φ, ψ) =
    \begin{cases}
      ψ, &\text{if }reorder_{∨}(φ, ψ) ≡ ψ\\
      ψ, &\text{if }ψ ≡ ψ₁ ∧ ψ₂, conjunct_{∨}(φ, ψ₁) ≡ ψ₁\\
        &\text{and } reorder_{∨}(φ, ψ₂) ≡ ψ₂\\
      ψ, &\text{if }φ ≡ φ₁ ∧ φ₂, conjunct_{∨}(φ₁, ψ) ≡ ψ\\
      ψ, &\text{if }φ ≡ φ₁ ∧ φ₂, conjunct_{∨}(φ₂, ψ) ≡ ψ\\
      φ, &\text{otherwise.}
    \end{cases}
  \end{split}
  \end{align*}
\end{lemma}

@@

In a disjunction, $φ = φ₁ ∨ φ₂ ∨ \cdots ∨ φₙ$, we say $ψ ∈_{∨} φ$,
if there is some $i = 1, \cdots, n$ such that $ψ = φᵢ$.
Then, we define the $rm_{∨}$ function to remove in a disjunction φ, those propositions ψ repeated in the formula. The formula $φ$ must be a right-associative disjunction.

@@

\begin{equation*}
\label{eq:rm-and-or}
  rm_{∧∨}(φ) =
  \begin{cases}
    rm_{∧∨}(φ₁) ∧ rm_{∧∨}(φ₂), &\text{if }φ ≡ φ₁ ∧ φ₂\\
    rm_{∨}(rdisj(φ)),  &\text{otherwise.}
  \end{cases}
\end{equation*}

@@

Finally, we remove redundancies in a proposition assuming that it is given in conjunctive normal form and right-associative, by composing the functions described above in the Eq.~\ref{eq:redun-ast}
for the function $redun^{*}$.

\begin{equation*}
  \label{eq:redun-ast}
  redun^{*} = rm_{∧} ∘ rm_{∧∨}.
\end{equation*}

Then, performing rearrangements to match with the expected output
by $redon^{*}$ function.

\begin{equation*}
  \label{eq:redun}
  redun(φ) = reorder_{∧∨}(φ, redun^{*}(φ)).
\end{equation*}

\begin{theorem}[thm-redun]
  \label{thm:thm_redun}
  $Γ ⊢ φ$ then $Γ ⊢ redun(φ)$.
\end{theorem}

\begin{proof}
Use in Theorem~\ref{thm:thm-reorder-and-or} with $ψ$ equals to
$redun^{*}(φ)$.
\end{proof}
